---
title: "ENVS 193DS Final"
author: "Helena Parsons"
format: pdf
editor: visual
---

# Link to Github repo

https://github.com/hparsons333/ENVS-193DS_spring-2026_final

# Setting up

```{r}
#| label: loading-packages
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(MuMIn)
library(ggeffects)
library(DHARMa)
```

# Problem 1. Research writing

**You’re working on a research team trying to understand how restoration of native plant cover (measured in percent cover) influences probability of American Avocet (Recurvirostra americana) and other shorebird habitat use in the San Francisco Bay. Your co-worker runs some analyses and writes up a report, giving it to you to review. In part 1 of the results section of the report, your co-worker has written:**

***We rejected the null hypothesis that distance to water edge does not predict probability of American Avocet microhabitat use (p = 0.03).***

**In part 2 of the results section of the report, your co-worker has written:**

***We rejected the null hypothesis that there is no relationship between bird species (American Avocet, Forster’s Tern, Black-necked Stilt) and habitat (managed pond, non-tidal marsh, salt pond, tidal marsh, or tidal mudflat) (p \< 0.001).***

**It’s great that this report is coming together, but you think they can improve on what they’ve written and make it more understandable to a non-statistical audience.**

## 1a. Transparent statistical methods

**What statistical tests did your co-worker use?**

**Clearly connect the test to the part that you are addressing (e.g. “In part 1, they used ________. In part 2, they used ________.”).**

In part 1, they used a simple linear regression to test if distance to water edge predicted American Avocet microhabitat use. In part 2, they used a correlation coefficient test to see if there was a relationship between bird species and habitat.

## 1b. Suggestions for rewriting

**In 1-3 sentences, write new statements with interpretation only for the statements in part a and b.**

**Be sure your interpretation makes sense without any statistical summary (which, again, you should not include).**

### Rewriting statement a

Our results suggest that we can reject the null hypothesis that distance to water edge does not predict probability of American Avocet microhabitat use. Alternatively, we determine that the distance to water edge significantly predicts whether or not American Avocent will utilize microhabitats.

### Rewriting statement b.

Our results suggest that we can reject the null hypothesis, which states that there is no relationship between the type of bird species found and the habitat type. Alternatively, we determine that there is a significant relationship between the type of bird species found and the habitat type.

## 1c. Further analysis

In 1-2 sentences, outline:

**- one additional variable that could influence the probability of American Avocet microhabitat use**

**- what type of variable it is (with units and/or levels, if applicable)**

**- why that variable might influence American Avocet habitat use.**

**Be specific with units, data collection method, etc. as needed.**

Because American Avocet need their nesting, brooding-rearing, foraging, and roosting sites to be protected from predators, another variable that could influence the probability of American Avocet habitat use is island distance from landmass/shore, since increased distance from the nearest landmass would provide increased distance from larger predators.

This would be a continuous variable, measured in distance in kilometers (since the area of the pond is measured in kilometers squared).

# Problem 2. Data analysis

## Reading in data

```{r}
#| label: reading-in-data
#| message: false
#| warning: false

# creating new object
nests <- 
  # reading in csv
  read_csv(
    # directing function to data folder
    here("data", 
         # csv: nest_data_final
         "nest_data_final.csv"))
```

## 2a. Clean and wrangle

```{r}
#| label: cleaning-and-wrangling-data

# creating new object
nests_clean <- 
  # df: nests
  nests |>
  # selecting only necessary columns: tree height, bird nest occurrence, and ant species present
  select(height_cm, case_control, ant_species) |>
  # filtering ant_species column to not include NA or T. penzigi
  filter(ant_species == "RRB" | ant_species == "AB" | ant_species == "BBR") |>
  # creating new column to store ant species' binomial names
  mutate(ant_species_name = case_match(
    ant_species,
    "RRB" ~ "Crematogaster mimosae", # C. mimosae
    "AB" ~ "Crematogaster sjostedti", # C. sjostedti
    "BBR" ~ "Crematogaster nigriceps")) |> # C. nigriceps
  # reordering ant species levels
  mutate(ant_species_name = as_factor(ant_species_name),
         ant_species_name = fct_relevel(ant_species_name,
                                    "Crematogaster mimosae",
                                    "Crematogaster sjostedti",
                                    "Crematogaster nigriceps"))

  
# creating new object
  nests_clean_display <- 
    # df: nests_clean
    nests_clean |>
    # selecting 10 random observations
    slice_sample(n = 10)
  
  # displaying object
  nests_clean_display
  
  str(nests_clean)
```

## 2b. Response variable

**In 1-2 sentences, explain what the 1s and 0s mean in this data set biologically.**

A 1 value in the case_control column corresponds to a tree that was observed to have a birds nest. A 0 value corresponds to a tree that did not have a birds nest, or was "available" for one.

## 2c. Purpose of study

**In 2-3 sentences for each predictor, describe:**

**- what kind of variable it is (i.e. continuous, discrete, categorical, etc.)** **- how it could possibly influence the probability of nest occurrence (i.e. as tree height increases, the probability of probability of nest occurrence would \_\_\_\_\_\_\_\_\_\_\_\_\_\_ because…)** **- where in the paper you found this information (the specific section, subsection, and paragraph)**

The type of ant species found is a categorical variable. Nest occurrence is predicted to be influenced by the aggression of the ant species found in each tree, and nest occurrence will increase with more aggressive species of ants found, because ants will indirectly protect the nests from predation. I found information about the hypothesis in fourth paragraph of the introduction, the collection of the data for ants in the first paragraph of the methods. and the results for the model of ant species as a predictor in the third paragraph of the results.

Tree height is continuous data. Nest occurrence may be influenced by tree height in that as tree height increases, nest occurrence probability will increase, as being further from the ground provides more protection from predators. I found the information about how tree height data was collected in the first paragraph of the methods, the results for the model of tree height as a predictor in the second paragraph of the results section, and information on what type of data it was in the table of the results section.

## 2d. Table of models

| Model number | Ant species | Tree height | Predictor list |
|-------------|:-----------:|:-----------:|:-------------------------------:|
| 0 | X | X | all predictors (full model) |
| 1 |  | x | tree height W/interaction effect of ant species |

## 2e. Run the models

```{r}

# model 0: all predictors
model0 <- lm(
  case_control ~ ant_species_name + height_cm, # formula: all predictors
  data = nests_clean) # clean data frame

# model 1: tree height as a predictor with interaction effect of ant_species
model1 <- lm(
  case_control ~ height_cm * ant_species_name, # formula: height_cm w/interaction effect of ant_species
  data = nests_clean) # clean data frame
```

## 2f. Select the best model

```{r}

# calculating lowest AIC value between models
AICc(model0,
     model1)
```

The best model as determined by Akaike’s Information Criterion (AIC) was the model where acacia ant species identity influenced the effect of tree height on probability of nest occurrence.

## 2g. Check diagnostics

```{r}
# displaying plots in 2x2 grid
par(mfrow = c(2,2))
# generating diagnostic plots
plot(model1)
```

## 2h. Visualize the model predictions

```{r}
#| label: model-predictions

# creating a new object
preds <- 
  # generating model based estimate
  ggpredict(
    # model: model1
    model1,
    # terms: case_control as predicted by effect of height_cm by ant species
                   terms = c("height_cm", "ant_species_name")) |>
  # renaming
  rename(height_cm = x,
         ant_species_name = group)
```

```{r}
#| label: visualizing-model

# base layer: ggplot
ggplot(
  # df: clean nests data
  data = nests_clean,
  # x-axis: tree height
  # y-axis: bird nest occurrence 
       mapping = aes(x = height_cm,
                     y = case_control,
                     # color by ant species
                     color = ant_species_name)) +
  # first layer: jitterplot
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21,
              alpha = 0.8) +
  # second layer: model estimate
  geom_pointrange(data = preds,
                  mapping = aes(x = height_cm,
                                y = predicted,
                                ymin = conf.low,
                                ymax = conf.high)) +
  scale_color_manual(values = c("darkorchid", "turquoise4", "hotpink")) +
  labs(title = "Effect of tree height on nest occurrence differs by acacia ant species",
       x = "Tree height (cm)",
       y = "Nest presence") + 
  theme_bw() +
  theme(legend.position = "none") +
  theme(panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()) +
  facet_wrap(~ ant_species_name)
```

## 2i. Write a caption for your figure

**Figure 1. Effect of tree height on nest occurrence differs by acacia ant species.** Small transparent dots represent measurements of nest occurrence at different tree heights in the presence of three different species of ants: C. mimosae, C. sjostedti, and C. nigriceps. Large filled points represent estimated nest occurrence at different tree heights in the presence of the three ant species. Lujan, E., Nielsen, R., Short, Z., Wicks, S., Nderitu Watetu, W., Malingati Khasoha, L., Palmer, T. M., Goheen, J. R., & Alston, J. M. (2023). Data, code, and metadata for: Symbiotic acacia ants drive nesting behavior by birds in an African savanna \[Data set\]. Zenodo. https://doi.org/10.5281/zenodo.8373322

## 2j. Calculate model predictions

```{r}
#| label: model-predictions

# creating a new object
preds_2 <- 
  # generating model based estimate
  ggpredict(
    # model: model1
    model1,
    # terms: case_control as predicted by effect of height_cm by ant species
                   terms = c("height_cm", "ant_species_name")) |>
  # renaming
  rename(height_cm = x,
         ant_species_name = group)
```

## 2k. Interpret your results

# Problem 3. Affective and explaoratory visualizations

## 3a. Comparing visualizations

**Compare and contrast your affective visualization from Homework 3 and the exploratory visualizations you made for Homework 2. In 1-3 sentences each, explain:**

**- How are the visualizations different from each other in the way you have represented your data?**

The visualization that I created in homework 2 were very basic, and didn't clearly display a trend or the relationship I was trying to visualize (although this may have also been because I didn't have enough data points yet). The affective visualization was a lot more abstract, and did require the artist statement to be fully understood, but I feel like it represented the relationship I wanted to display from the figures that I choose to represent the data point and the background I chose as well.

**- What similarities do you see between all your visualizations?**

I chose to use the scatter plot for both my exploratory visualization and my affective visualization, and between all my visualizations there seems to be the most weight on the left side of the axis rather than the right, most likely because both the stress and outside engagement variables increased as time spent on the job search decreased.

**- What patterns (e.g. differences in means/counts/proportions/medians, trends through time, relationships between variables) do you see in each visualization? Are these different between visualizations? If so, why? If not, why not?** 

I mostly see what could be described as a negative linear relationship, or a negative trend, between both my outside engagements and time spent on the job search, and my level of stress and time spent on the job search. These are similar because when I am stressed or when I have a lot of outside engagements, I either feel less motivated to spend time on a job search (because I am concerned about other work that I have) or because I do not have the time available to apply. 

**- What kinds of feedback did you get during week 9 in workshop or from the instructors? How did you implement or try those suggestions? If you tried and kept those suggestions, explain how and why; if not, explain why not.**

The two suggestions that I got from workshop and implemented were to use briefcases as representations of my data points, and to color in the balloons to represent decreasing and increasing levels of stress. These were both extremely helpful, as using briefcases was more effective at representing the response variable I had chosen, and coloring the balloons by stress allowed me to get rid of visual clutter (where I initially created a color gradient below the data points) and represent increasing feelings of stress through the transition of cool colors to warm colors.

##
