---
title: "ENVS 193DS Final"
author: "Helena Parsons"
format: pdf
editor: visual
---

# Link to Github repo

https://github.com/hparsons333/ENVS-193DS_spring-2026_final

# Setting up

```{r}
#| label: loading-packages
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(MuMIn)
library(ggeffects)
library(DHARMa)
```

# Problem 1. Research writing

**You’re working on a research team trying to understand how restoration of native plant cover (measured in percent cover) influences probability of American Avocet (Recurvirostra americana) and other shorebird habitat use in the San Francisco Bay. Your co-worker runs some analyses and writes up a report, giving it to you to review. In part 1 of the results section of the report, your co-worker has written:**

***We rejected the null hypothesis that distance to water edge does not predict probability of American Avocet microhabitat use (p = 0.03).***

**In part 2 of the results section of the report, your co-worker has written:**

***We rejected the null hypothesis that there is no relationship between bird species (American Avocet, Forster’s Tern, Black-necked Stilt) and habitat (managed pond, non-tidal marsh, salt pond, tidal marsh, or tidal mudflat) (p \< 0.001).***

**It’s great that this report is coming together, but you think they can improve on what they’ve written and make it more understandable to a non-statistical audience.**

## 1a. Transparent statistical methods

**What statistical tests did your co-worker use?**

**Clearly connect the test to the part that you are addressing (e.g. “In part 1, they used \_\_\_\_\_\_\_. In part 2, they used \_\_\_\_\_\_\_.”).**

In part 1, they used a simple linear regression to test if distance to water edge predicted American Avocet microhabitat use. In part 2, they used a correlation coefficient test to see if there was a relationship between bird species and habitat.

## 1b. Suggestions for rewriting

**In 1-3 sentences, write new statements with interpretation only for the statements in part a and b.**

**Be sure your interpretation makes sense without any statistical summary (which, again, you should not include).**

### Rewriting statement a

Our results suggest that we can reject the null hypothesis that distance to water edge does not predict probability of American Avocet microhabitat use. Alternatively, we determine that the distance to water edge significantly predicts whether or not American Avocent will utilize microhabitats.

### Rewriting statement b.

Our results suggest that we can reject the null hypothesis, which states that there is no relationship between the type of bird species found and the habitat type. Alternatively, we determine that there is a significant relationship between the type of bird species found and the habitat type.

## 1c. Further analysis

In 1-2 sentences, outline:

**- one additional variable that could influence the probability of American Avocet microhabitat use**

**- what type of variable it is (with units and/or levels, if applicable)**

**- why that variable might influence American Avocet habitat use.**

**Be specific with units, data collection method, etc. as needed.**

Because American Avocet need their nesting, brooding-rearing, foraging, and roosting sites to be protected from predators, another variable that could influence the probability of American Avocet habitat use is island distance from landmass/shore, since increased distance from the nearest landmass would provide increased distance from larger predators.

This would be a continuous variable, measured in distance in kilometers (since the area of the pond is measured in kilometers squared).

# Problem 2. Data analysis

## Reading in data

```{r}
#| label: reading-in-data
#| message: false
#| warning: false

# creating new object
nests <- 
  # reading in csv
  read_csv(
    # directing function to data folder
    here("data", 
         # csv: nest_data_final
         "nest_data_final.csv"))
```

## 2a. Clean and wrangle

```{r}
#| label: cleaning-and-wrangling-data

# creating new object
nests_clean <- 
  # df: nests
  nests |>
  # selecting only necessary columns: tree height, bird nest occurrence, and ant species present
  select(height_cm, case_control, ant_species) |>
  # filtering ant_species column to not include NA or T. penzigi
  filter(ant_species == "RRB" | ant_species == "AB" | ant_species == "BBR") |>
  # reordering ant species levels
  mutate(ant_species = as_factor(ant_species),
         ant_species = fct_relevel(ant_species,
                                    "RRB",
                                    "AB",
                                    "BBR")) |>
   # creating new column to store ant species' binomial names
  mutate(ant_species_name = case_match(
    ant_species,
    "RRB" ~ "Crematogaster mimosae", # C. mimosae
    "AB" ~ "Crematogaster sjostedti", # C. sjostedti
    "BBR" ~ "Crematogaster nigriceps")) # C. nigriceps
  
# creating new object
  nests_clean_display <- 
    # df: nests_clean
    nests_clean |>
    # selecting 10 random observations
    slice_sample(n = 10)
  
  # displaying object
  nests_clean_display
  
  str(nests_clean)
```

## 2b. Response variable

**In 1-2 sentences, explain what the 1s and 0s mean in this data set biologically.**

A 1 value in the case_control column corresponds to a tree that was observed to have a birds nest. A 0 value corresponds to a tree that did not have a birds nest, or was "available" for one. 

## 2c. Purpose of study

**In 2-3 sentences for each predictor, describe:**

**- what kind of variable it is (i.e. continuous, discrete, categorical, etc.)**
**- how it could possibly influence the probability of nest occurrence (i.e. as tree height increases, the probability of probability of nest occurrence would ______________ because…)**
**- where in the paper you found this information (the specific section, subsection, and paragraph)**

The type of ant species found is a categorical variable. Nest occurrence is predicted to be influenced by the aggression of the ant species found in each tree, and nest occurrence will increase with more aggressive species of ants found, because ants will indirectly protect the nests from predation. I found information about the hypothesis in fourth paragraph of the introduction, the collection of the data for ants in the first paragraph of the methods. and the results for the model of ant species as a predictor in the third paragraph of the results.

Tree height is continuous data. Nest occurrence may be influenced by tree height in that as tree height increases, nest occurrence probability will increase, as being further from the ground provides more protection from predators. I found the information about how tree height data was collected in the first paragraph of the methods, the results for the model of tree height as a predictor in the second paragraph of the results section, and information on what type of data it was in the table of the results section.

## 2d. Table of models

| Model number | Ant species | Tree height | Predictor list                                 |      
|--------------|:-----------:|:-----------:|:----------------------------------------------:| 
| 0            |      X      |     X       | all predictors (full model)                    |
| 1            |             |     x       | tree height W/interaction effect of ant species|

## 2e. Run the models

```{r}

# model 0: all predictors
model0 <- lm(
  case_control ~ ant_species + height_cm, # formula: all predictors
  data = nests_clean) # clean data frame

# model 1: tree height as a predictor with interaction effect

model1 <- lm(
  case_control ~ height_cm * ant_species, # formula: all predictors
  data = nests_clean) # clean data frame
```

## 2f. Select the best model    

```{r}
AICc(model0,
     model1)
```

The best model as determined by Akaike’s Information Criterion (AIC) was the model where acacia ant species identity influenced the effect of tree height on probability of nest occurrence.

## 2g. Check diagnostics
